Fluid-mechanical erosion creates anisotropic porous media

### 2022-05-26 docs/PhysicsPaper/StokesErosion.tex
NM

% New
\newcommand{\anis}{\mathcal{A}}

@@

While the Stokes equations \eqref{eqn:StokesEq} provides a microscopic description of the detailed flow field $\uu$ penetrating the complex configuration of erodable bodies, a coarse-grained description can be obtained by treating the collection of bodies as a single porous-medium and homogenizing the flow-field through Darcy's law,
\begin{equation}
\bq = - \bvec{k} \grad p / \mu
\end{equation}
Here, $p$ and $\mu$ represent the pressure field and fluid viscosity as before, with $\mu = 1$ by the non-dimensionalization. Meanwhile, $\bq$ represents the {\em specific discharge}, which is the volume of water that flowing through a unit cross sectional area of porous media per unit time; $\bq$ relates to the (interstitial) velocity $\uu$, by integrating $\uu$ over a sufficiently small control region and dividing by the total volume (including both fluid and solid) of the region. The parameter $\bvec{k}$ represents the {\em permeability} of the porous medium, which generally takes the form of a rank-2 tensor to permit different propensities to flow in different directions, i.e.~medium anisotropy. The permeability may vary in space and will certainly vary in time as the bodies erode.

Consider, first for simplicity, an isotropic porous medium with scalar permeability $k$.

@@

@@

Consider first an isotropic porous medium with {\em scalar} permeability $k$. Furthermore, we seek to characterize the medium by a single, bulk value permeability, so we assume that $k$ is spatially constant.
Equivalently, we can introduce a {\em resistivity} simply defined as the reciprocal of permeability, $r = 1/k$. Then taking the horizontal component gives
\begin{equation}
\label{q1}
q_1(x,y) = -\frac{1}{r} \pderiv{p}{x}(x,y)
\end{equation}
Consider the horizontal discharge, $q_1(x,y)$, averaged over a vertical slice
\begin{equation}
\qavg := \frac{1}{2} \int_{-1}^{1} q_1(x_0, y) dy = U
\end{equation}
By conservation of mass, $\qavg$ is independent of the horizontal position $x_0$ and is equal to the imposed value $U$. Since $\bq$ is a volume-average of the tracer velocity, $\qavg$ can be equivalently obtained by averaging the tracer velocity, which is how we will measure this quantity in the simulations.

Next, consider the pressure averaged over a vertical slice
\begin{equation}
\pavg(x_0) := \frac{1}{2} \int_{-1}^{1} p(x_0, y) dy
\end{equation}
In particular, the upstream and downstream pressures are respectively
\begin{equation}
\pup = \pavg(-1) \, , \quad \pdn = \pavg(1)
\end{equation}

Integrating \eqref{q1} over the porous-medium domain $[-1, 1] \times [-1,1]$, applying the fundamental theorem of calculus, and rearranging gives
\begin{equation}
r = \frac{\pup - \pdn}{2 U}
\end{equation}
This exact formula gives the resistivity based on measurements of the upstream/downstream pressure and the discharge.

Probably should just use
\begin{equation}
k = \frac{2U}{\pup - \pdn}
\end{equation}
%Note: If a larger value of $x_0$ is chosen, then one would be in effect averaging the porous-medium resistivity with the vanishing resistivity of the empty buffer region.

>>>>

% New
\newcommand{\anis}{\mathcal{A}}
\newcommand{\diag}{\mathop{\mathrm{diag}}}

@@

While the Stokes equations \eqref{eqn:StokesEq} provides a microscopic description of the detailed flow field $\uu$ penetrating the complex configuration of erodable bodies, a coarse-grained description can be obtained by treating the collection of bodies as a single porous-medium and homogenizing the flow-field through Darcy's law,
\begin{equation}
\label{eqn:Darcy}
\bq = - \frac{1}{\mu} \bvec{k} \grad p
\end{equation}
Here, $p$ and $\mu$ represent the pressure field and fluid viscosity as before, with $\mu = 1$ by the non-dimensionalization. Meanwhile, $\bq = (q_1, q_2)$ represents the {\em specific discharge}, which is the volume of water that flowing through a unit cross sectional area of porous media per unit time; $\bq$ relates to the (interstitial) velocity $\uu$, by integrating $\uu$ over a sufficiently small control region and dividing by the total volume (including both fluid and solid) of the region. The parameter $\bvec{k}$ represents the {\em permeability} of the porous medium, which generally takes the form of a rank-2 tensor to permit different propensities to flow in different directions, i.e.~medium anisotropy.

%The permeability may vary in space and will certainly vary in time as the bodies erode.

@@

For simplicity, we will assume $\bvec{k}  = \diag(k_{11}, k_{22})$ to be a diagonal matrix for the sake of characterizing the permeability of the porous medium. Because the diagonal components of $\bvec{k}$ need not be equal, the medium can have different permeabilities in the horizontal and vertical directions. Further, we will assume that $\bvec{k}$ to be spatially homogeneous in order to characterize the medium with a single bulk quantity at any instance in time. Naturally, the permeability will change over time as the bodies that comprise the medium erode. Consider first the horizontal component of \eqref{eqn:Darcy}
\begin{equation}
\label{q1}
q_1(x,y) = -k_{11} \pderiv{p}{x}(x,y)
\end{equation}

By conservations of mass, the average of the horizontal discharge, $q_1$, over any vertical cross-section must equal the uniform flow rate $U$ imposed at the inlet/outlet. That is, for any location $x_0$,
\begin{equation}
\qavg_1 := \frac{1}{2} \int_{-1}^{1} q_1(x_0, y) dy = U
\end{equation}
Above and henceforth, the overline signifies an average over a vertical cross-section. Similarly, consider the pressure averaged over a vertical cross-section at $x_0$
\begin{equation}
\pavg(x_0) := \frac{1}{2} \int_{-1}^{1} p(x_0, y) dy
\end{equation}
In particular, we define the {\em upstream} and {\em downstream} pressures as
\begin{equation}
\pup = \pavg(-1) \, , \quad \pdn = \pavg(1)
\end{equation}
since $x_0 = -1$ lies immediately upstream of the porous medium and $x_0 = 1$ immediately downstream.

Integrating \eqref{q1} over the porous-medium domain $[-1, 1] \times [-1,1]$, applying the fundamental theorem of calculus, and rearranging gives
\begin{equation}
k_{11} = \frac{2U}{\pup - \pdn}
\end{equation}
This exact formula gives the permeability in terms of the total flux $U$ and measurements of the upstream/downstream pressure.

@@

%Since $\bq$ is a volume-average of the tracer velocity, $\qavg$ can be equivalently obtained by averaging the tracer velocity, which is how we will measure this quantity in the simulations.

%Note: If a larger value of $x_0$ is chosen, then one would be in effect averaging the porous-medium resistivity with the vanishing resistivity of the empty buffer region.
